The Unicode Cookbook \linebreak for Linguists

### 2018-06-08 book/chapters/ipa_background.tex
through paperhivers chapter 4

Essentially, transcription has two parts: a text containing symbols and a set 
of conventions for interpreting those symbols (and their combinations). 
The symbols of the IPA distinguish between letter-like symbols and
diacritics (symbol modifiers). The use of the letter-like symbols to represent 
a language's sounds is guided by the principle of contrast; where two words 
are distinguishable by phonemic contrast, those contrasts should be transcribed 
with different letter symbols (and not just diacritics). Allophonic distinction 
falls under the rubric of diacritically-distinguished symbols, e.g.\ the 
difference in English between an aspirated /p/ and /t/ in [spʰot̚] and 
an unreleased /p/ in [stʰop̚].

@@

There are different reasons for this difficulty to interpret IPA symbols, all
officially sanctioned by the IPA. An important principle of the IPA is that
different representations resulting from different underlying analyses are
allowed. Because the IPA does not provide phonological analyses for specific
languages, the IPA does not define a single correct transcription system.
Rather, the IPA aims to provide a resource that allows users to express any
analysis so that it is widely understood. Basically, the IPA allows for both a 
\textit{narrow} phonetic transcription and a \textit{broad} phonological transcription. 
A narrow phonetic transcription may freely use the all symbols in the IPA 
chart with direct reference to their phonetic value, i.e.\ the transcriber can 
indicate with the symbols <ŋ͡m> that the phonetic value of the attested sound 
is a simultaneous labial and velar closure which is voiced and contains nasal 
airflow, independently of the phonemic status of this sound in the language in 
question.

@@

In contrast, the basic goal of a phonemic broad transcription is to distinguish all
words in a language with a minimal number of transcription symbols
\citep[19]{Abercrombie1964}. A phonemic transcription includes the conventions
of a particular language's phonological rules. These rules determine the
realization of that language's phonemes. For example, in the transcription of
German, Dutch, English and French a symbol <t> might be used for the voiceless
plosive in the alveolar and dental areas. This symbol is sufficient for a succinct
transcription of these languages because there is no further phonemic
subdivisions in this domain in either of these languages. However, the
language-specific realization of this consonant is closer to [t̪ʰ], [t], [tʰ]
and [t̪], respectively. Similarly, the five vowels of Greek can be represented
phonemically in IPA as /ieaou/, though phonetically they are closer to [iεaɔu].
The Japanese five vowel system can also be transcribed in IPA as
/ieaou/, while the phonetic targets in this case are closer to [ieaoɯ].

@@

\subsection*{IPA Numbers}
Prior to the Kiel Convention for the modern revision of the IPA in 1989,
\citet{Wells1987} collected and published practical approaches to coded
representations of the IPA, which dealt mainly with the assignment of characters
on the keyboard to IPA symbols. The process of assigning standardized computer
codes to phonetic symbols was given to the \textsc{workgroup on computer
coding} (henceforth \textit{working group}) at the Kiel Convention. This working
group was assigned to consider the following tasks
\citep{Esling1990,EslingGaylord1993}:

@@

\noindent The identification of IPA symbols with unique identifiers was 
a first step in formalizing the IPA computationally because it would give 
each symbol an unambiguous numerical identifier called an \textsc{IPA number}. 
The numbering system was to be comprehensive enough to support future revisions 
of the IPA, including symbol specifications and diacritic placement. The 
application of diacritics was also to be made explicit.

Although the Association had never officially approved a set of names 
for the IPA symbols, each IPA symbol received a unique \textsc{IPA name}. 
Many symbols already had an informal name (or two) used by linguists, but 
consensus on symbol names was growing due to the recent publication of the 
\textit{Phonetic Symbol Guide} \citep{PullumLadusaw1986}. Thus most of the 
IPA symbol names were taken from that source \citep[31]{IPA1999}.

The working group decided insightfully that the computing-coding convention 
for the IPA should be independent of computer environments or formats, 
i.e.\ the IPA Number was not meant to be encoded at the bit pattern level.
The working group report's declaration includes the following explanatory 
remarks \citep[82]{International1989report}:

@@

\noindent The coding convention was not meant to address the engineering 
aspects of the actual encoding in computers (cf.\ \cite{Anderson1984}). However, 
it was meant to serve as a basis for a communication-interchange standard for 
creating mapping tables from various computer encodings, fonts, phonetic-character-set 
software, etc., to common IPA Numbers, and therefore symbols.\footnote{Remember, at 
this time in the late 1980s there was no stable multilingual computing environment. 
But some solution was needed because scholars were increasingly using personal 
computers for their research and many were quickly adopting electronic mail or 
discussion boards like Usenet as a medium for international exchanges. 
Most of these systems ran on 8-bit hardware systems using a 7-bit ASCII character encoding.}

@@

The working group assigned each IPA symbol to a unique three-digit number, i.e.\
an IPA Number. Encoded in this number scheme implicitly includes information
about the status of each symbol (see below). The IPA numbers were listed with the
IPA symbols and they were also illustrated in IPA chart form (see
\cite[84]{EslingGaylord1993} or \cite[App. 2]{IPA1999}). The numbers were
assigned in linear order (e.g.\ [p] 101, [b] 102, [t] 103...) following the IPA
revision of 1989 and its update in 1996. Although the numbering scheme still 
exists, in practice is it superseded by the Unicode codification of symbols.

The working group made the decision that no IPA symbol, past or present, 
could be ignored. The comprehensive inclusion of all IPA symbols was to 
anticipate the possibility that some symbols might be added, withdrawn, 
or reintroduced into current or future usage. For example, in the 1989 
revision voiceless implosives <~ƥ,~ƭ,~ƈ,~ƙ,~ʠ~> were added; in the 1993 
revision they were removed. Ligatures like <~ʧ,~ʤ~> are included as formerly 
recognized IPA symbols; they are assigned to the 200 series of IPA numbers 
as members of the group of symbols formerly recognized by the IPA. To ensure 
backwards compatibility, legacy IPA symbols would retain an IPA Number and 
an IPA Name for reference purposes. As we discuss below, this decision is 
later reflected in the Unicode Standard as many legacy IPA symbols still reside in 
the \textsc{IPA extensions} block.

The IPA Number is expressed as a three-digit number. The first digit 
indicates the symbol's category \citep{Esling1990,EslingGaylord1993}:

@@

After a symbol is categorized, it is assigned a number sequentially, e.g.\ [i]
301, [e] 302, [ɛ] 303. The system allows for the addition of new symbols within
the various series by appending them, e.g.\ [\charis{ⱱ}] 184. Former non-IPA
symbols (or often-used but non-official IPA symbols) for consonants, vowels and
diacritics are numbered from x99 backwards. For example, the voiceless and
voiced postalveolar affricates and fricatives <~č,~ǰ,~š,~ž~> are assigned the
IPA numbers 299, 298, 297 and 296, respectively, because they are not sanctioned
IPA symbols.

The assignment of the IPA numbers to IPA symbols provided the basis for uniquely
identifying the set of past and present IPA symbols as a type of computational
representational standard of the IPA. Within each revision of the IPA, the
coding defines a closed and clearly defined set of characters. The benefits of
this standardization are clear in at least two ways: it is used in translation
tables that reference ASCII representations of the IPA, and this early
computational representation of the IPA became the basis X-SAMPA and for the
inclusion of the IPA into the Unicode Standard version 1.0.

@@

True to the working group's aim, the IPA numbers provided a mechanism for 
a communication interchange standard for creating mapping tables to various 
computer encodings. For example, the IPA coding system was used as a mapping 
system in the creation of SAMPA \citep{Wells_etal1992}, an ASCII representation 
of the IPA symbols.

For a long time, linguists, like all other computer users, were
limited to ASCII-encoded 7-bit characters, which only includes Latin characters,
numbers and some punctuation and symbols. Restricted to these standard character
sets that lacked IPA support or other language-specific graphemes that they
needed, linguists devised their own solutions.\footnote{Early work addressing
the need for a universal computing environment for writing systems and their
computational complexity is discussed in \citet{Simons1989}. A more recent survey of
practical recommendations for language resources, including notes on encoding,
can be found in \citet{BirdSimons2003}.} For example, some chose to represent
unavailable graphemes with substitutes, e.g.~the combination of <ng> to
represent <ŋ>. Tech-savvy linguists redefined selected characters from a
character encoding by mapping custom made fonts to specific code points.\footnote{For 
example, SIL's popular font SIL IPA 1990.}. However,
one linguist's electronic text would not render properly on another linguist's
computer without access to the same font. Furthermore, if two character encodings
defined two character sets differently, then data could not be reliably and
correctly displayed. This is a commonly encountered example of the non-interoperability of
data and data formats.

One solution was the ASCII-ification of the IPA, which simply involved 
defining keyboard-able sequences consisting of ASCII combinations as IPA symbols. 
For example, \cite{Wells1987} provides an in-depth description of IPA
codings from country-to-country. Later ASCII-IPAs include Kirshenbaum (created
in 1992 in a Usenet group and named after its lead developer who was at
Hewlett-Packard Laboratories) and Worldbet (published in 1993 by
\cite{Hieronymus1993}, who was at AT\&T Laboratories). 
The most successful effort was SAMPA (Speech Assessment
Methods Phonetic Alphabet), which was created between 1988--1991 in Europe to 
represent IPA symbols with ASCII
character sequences \citep{Wells1987,Wells_etal1992}, using e.g.\ <p\textbackslash> 
for [ɸ]. SAMPA was developed by a group of speech scientists from nine countries 
in Europe and it constituted the ASCII-IPA symbols needed for phonemic transcription 
of the principal European languages \citep{Wells1995}. It is still widely 
used in language technology.

@@

During the 1980's, it became increasingly clear that an adequate solution 
to the problem of multilingual computing environments was needed. Linguists 
were on the forefront of addressing this issue because they faced these 
challenges head-on by wishing to publish and communicate electronic text 
with phonetic symbols which were not included in basic ASCII. One 
only needs to look at facsimiles of older electronic documents to see exotic 
symbols written in by hand after the preparation of typed version.

@@

After the Kiel Convention in 1989, the working group assisted
the International Phonetic Association in representing the IPA to the
\textsc{international organization for standardization} (ISO) and to the \textsc{text
encoding initiative} (TEI) \citep{EslingGaylord1993}. The working group's
formalization of the IPA, i.e.\ a full listing of agreed upon computer codings
for phonetic symbols, was used in developing writing system descriptions, which
were at the time being solicited for scripts to be included in the new
multilingual international character encoding standards. The working group for
ISO/IEC 10646 and Unicode were two such initiatives.

@@

Character set ISO/IEC 10646 was approved by ISO, including the phonetic
characters submitted to them in May 1993. The set of IPA characters were
assigned UCS codes in 16 bit representation (in hexadecimal) and were published
as Tables 2 and 3 in \cite{EslingGaylord1993}, which include a graphical
representation of the IPA Symbol, its IPA Name, phonetic description, IPA
Number, UCS Code and AFII Code.\footnote{The \textsc{association for font
information interchange} (AFII) was an international database of glyphs created
to promote the standardization of font data required to produce ISO/IEC 10646.} When the
character sets of ISO/IEC 10646 and the Unicode Standard later converged (see
Chapter~\ref{the-unicode-approach}), the IPA proposal was
included in the Unicode Standard Version 1.0 -- largely as we know it
today.

@@

It was a long journey, but the goal of achieving a single multilingual computing
environment has largely been accomplished. As such, we should not dismiss the IPA Numbers 
or pre-Unicode encoding attempts, such as SAMPA/X-SAMPA, as misguided. The parallels 
between the IPA Numbers and Unicode Code points, for example, is striking because both the IPA and 
the Unicode Consortium came up with the solution of an additional layer of indirection (an abstraction layer) 
between symbols/characters and encoding on the bit pattern level. SAMPA/X-SAMPA is also still useful 
as an input method for IPA in ASCII and required by some software.

@@

Unicode 88 provided the basic principles for the Unicode Standard's design -- 
pushing for 16 bit representations of characters with a clear distinction 
between characters and glyphs. Some of the contents of this status proposal 
of 1988 were reworked for inclusion in the early Unicode Standard pre-publication 
drafts and by August 1990, the proposal was in a (very) rough draft format. Its 
editors and the Unicode Working Group (the predecessor to the Unicode Technical 
Committee) worked together to lay out the proposed standard's structure and 
content. At this time, the proposal contained no code charts nor block descriptions.

>>>>

Essentially, transcription has two parts: a text containing symbols and a set 
of conventions for interpreting those symbols (and their combinations). 
The symbols of the IPA distinguish between letter-like symbols and
diacritics (symbol modifiers). The use of the letter-like symbols to represent 
a language's sounds is guided by the principle of contrast; where two words 
are distinguishable by phonemic contrast, those contrasts should be transcribed 
with different letter symbols (and not just diacritics). Allophonic distinction 
falls under the rubric of diacritically-distinguished symbols, e.g.\ the 
difference in English between an aspirated /p/ in [pʰæt] and 
an unreleased /p/ in [stop̚].

@@

There are different reasons for this difficulty to interpret IPA symbols, all
officially sanctioned by the IPA. An important principle of the IPA is that
different representations resulting from different underlying analyses are
allowed. Because the IPA does not provide phonological analyses for specific
languages, the IPA does not define a single correct transcription system.
Rather, the IPA aims to provide a resource that allows users to express any
analysis so that it is widely understood. Basically, the IPA allows for both a 
\textit{narrow} phonetic transcription and a \textit{broad} phonological transcription. 
A narrow phonetic transcription may freely use all symbols in the IPA 
chart with direct reference to their phonetic value, i.e.\ the transcriber can 
indicate with the symbols <ŋ͡m> that the phonetic value of the attested sound 
is a simultaneous labial and velar closure which is voiced and contains nasal 
airflow, independently of the phonemic status of this sound in the language in 
question.

@@

In contrast, the basic goal of a broad phonemic transcription is to distinguish all
words in a language with a minimal number of transcription symbols
\citep[19]{Abercrombie1964}. A phonemic transcription includes the conventions
of a particular language's phonological rules. These rules determine the
realization of that language's phonemes. For example, in the transcription of
German, Dutch, English and French a symbol <t> might be used for the voiceless
plosive in the alveolar and dental areas. This symbol is sufficient for a succinct
transcription of these languages because there is no further phonemic
subdivisions in this domain in either of these languages. However, the
language-specific realization of this consonant is closer to [t̪ʰ], [t], [tʰ]
and [t̪], respectively. Similarly, the five vowels of Greek can be represented
phonemically in IPA as /ieaou/, though phonetically they are closer to [iεaɔu].
The Japanese five-vowel system can also be transcribed in IPA as
/ieaou/, while the phonetic targets in this case are closer to [ieaoɯ].

@@

\subsection*{IPA Numbers}
Prior to the Kiel Convention for the modern revision of the IPA in 1989,
\citet{Wells1987} collected and published practical approaches to coded
representations of the IPA, which dealt mainly with the assignment of characters
on the keyboard to IPA symbols. The process of assigning standardized computer
codes to phonetic symbols was given to the \textit{workgroup on computer
coding} (henceforth \textit{working group}) at the Kiel Convention. This working
group was given the following tasks
\citep{Esling1990,EslingGaylord1993}:

@@

\noindent The identification of IPA symbols with unique identifiers was 
a first step in formalizing the IPA computationally because it would give 
each symbol an unambiguous numerical identifier called an \textsc{ipa number}. 
The numbering system was to be comprehensive enough to support future revisions 
of the IPA, including symbol specifications and diacritic placement. The 
application of diacritics was also to be made explicit.

Although the Association had never officially approved a set of names 
for the IPA symbols, each IPA symbol received a unique \textsc{ipa name}. 
Many symbols already had an informal name (or two) used by linguists, but 
consensus on symbol names was growing due to the recent publication of the 
\textit{Phonetic Symbol Guide} \citep{PullumLadusaw1986}. Thus most of the 
IPA symbol names were taken from that source \citep[31]{IPA1999}.

The working group decided insightfully that the computing-coding convention 
for the IPA should be independent of computer environments or formats, 
i.e.\ the \textsc{ipa number} was not meant to be encoded at the bit pattern level.
The working group report's declaration includes the following explanatory 
remarks \citep[82]{International1989report}:

@@

\noindent The coding convention was not meant to address the engineering 
aspects of the actual encoding in computers (cf.\ \cite{Anderson1984}). However, 
it was meant to serve as a basis for a interchange standard for 
creating mapping tables from various computer encodings, fonts, phonetic-character-set 
software, etc., to common IPA numbers, and therefore symbols.\footnote{Remember, at 
this time in the late 1980s there was no stable multilingual computing environment. 
But some solution was needed because scholars were increasingly using personal 
computers for their research and many were quickly adopting electronic mail or 
discussion boards like Usenet as a medium for international exchanges. 
Most of these systems ran on 8-bit hardware systems using a 7-bit ASCII character encoding.}

@@

The working group assigned each IPA symbol to a unique three-digit IPA number. 
Encoded in this number scheme implicitly is information
about the status of each symbol (see below). The IPA numbers were listed with the
IPA symbols and they were also illustrated in IPA chart form (see
\cite[84]{EslingGaylord1993} or \cite[App. 2]{IPA1999}). The numbers were
assigned in linear order (e.g.\ [p] 101, [b] 102, [t] 103...) following the IPA
revision of 1989 and its update in 1996. Although the numbering scheme still 
exists, in practice it is superseded by the Unicode codification of symbols.

The working group made the decision that no IPA symbol, past or present, 
could be ignored. The comprehensive inclusion of all IPA symbols was to 
anticipate the possibility that some symbols might be added, withdrawn, 
or reintroduced into current or future usage. For example, in the 1989 
revision voiceless implosives <~ƥ,~ƭ,~ƈ,~ƙ,~ʠ~> were added; in the 1993 
revision they were removed. Ligatures like <~ʧ,~ʤ~> are included as formerly 
recognized IPA symbols; they are assigned to the 200 series of IPA numbers 
as members of the group of symbols formerly recognized by the IPA. To ensure 
backwards compatibility, legacy IPA symbols would retain an IPA number and 
an IPA name for reference purposes. As we discuss below, this decision is 
later reflected in the Unicode Standard as many legacy IPA symbols still reside in 
the \textsc{IPA Extensions} block.

The IPA number is expressed as a three-digit number. The first digit 
indicates the symbol's category \citep{Esling1990,EslingGaylord1993}:

@@

After a symbol is categorized, it is assigned a number sequentially, e.g.\ [i]
301, [e] 302, [ɛ] 303. The system allows for the addition of new symbols within
the various series by appending them, e.g.\ [\charis{ⱱ}] 184. Former non-IPA
symbols (or often-used but non-official IPA symbols) for consonants, vowels and
diacritics are numbered from 299 backwards. For example, the voiceless and
voiced postalveolar affricates and fricatives <~č,~ǰ,~š,~ž~> are assigned the
IPA numbers 299, 298, 297 and 296, respectively, because they are not sanctioned
IPA symbols.

The assignment of the IPA numbers to IPA symbols provided the basis for uniquely
identifying the set of past and present IPA symbols as a type of computational
representational standard of the IPA. Within each revision of the IPA, the
coding defines a closed and clearly defined set of characters. The benefits of
this standardization are clear in at least two ways: it is used in translation
tables that reference ASCII representations of the IPA, and this early
computational representation of the IPA became the basis for X-SAMPA and for the
inclusion of the IPA into the Unicode Standard version 1.0.

@@

True to the working group's aim, the IPA numbers provided a mechanism for 
an interchange standard for creating mapping tables to various 
computer encodings. For example, the IPA coding system was used as a mapping 
system in the creation of SAMPA \citep{Wells_etal1992}, an ASCII representation 
of the IPA symbols.

For a long time, linguists, like all other computer users, were
limited to ASCII-encoded 7-bit characters, which only includes Latin characters,
numbers and some punctuation and symbols. Restricted to these standard character
sets that lacked IPA support or other language-specific graphemes that they
needed, linguists devised their own solutions.\footnote{Early work addressing
the need for a universal computing environment for writing systems and their
computational complexity is discussed in \citet{Simons1989}. A more recent survey of
practical recommendations for language resources, including notes on encoding,
can be found in \citet{BirdSimons2003}.} For example, some chose to represent
unavailable graphemes with substitutes, e.g.~the combination of <ng> to
represent <ŋ>. Tech-savvy linguists redefined selected characters from a
character encoding by mapping custom-made fonts to specific code points.\footnote{For 
example, SIL's popular font SIL IPA 1990.} However,
one linguist's electronic text would not render properly on another linguist's
computer without access to the same font. Furthermore, if two character encodings
defined two character sets differently, then data could not be reliably and
correctly displayed. This is a commonly encountered example of the non-interoperability of
data and data formats.

One solution was the ASCII-ification of the IPA, which simply involved 
defining keyboard-able sequences consisting of ASCII combinations as IPA symbols. 
For example, \citet{Wells1987} provides an in-depth description of IPA
codings from country-to-country. Later ASCII-IPAs include Kirshenbaum (created
in 1992 in a Usenet group and named after its lead developer who was at
Hewlett-Packard Laboratories) and Worldbet (published by
\citet{Hieronymus1993}, who was at AT\&T Laboratories). 
The most successful effort was SAMPA (Speech Assessment
Methods Phonetic Alphabet), which was created between 1988--1991 in Europe to 
represent IPA symbols with ASCII
character sequences \citep{Wells1987,Wells_etal1992}, using e.g.\ <p\textbackslash> 
for [ɸ]. SAMPA was developed by a group of speech scientists from nine countries 
in Europe and it constituted the ASCII-IPA symbols needed for phonemic transcription 
of the principal European languages \citep{Wells1995}. It is still widely 
used in language technology.

@@

During the 1980s, it became increasingly clear that an adequate solution 
to the problem of multilingual computing environments was needed. Linguists 
were on the forefront of addressing this issue because they faced these 
challenges head-on by wishing to publish and communicate electronic text 
with phonetic symbols which were not included in basic ASCII. One 
only needs to look at facsimiles of older electronic documents to see exotic 
symbols written in by hand after the preparation of typed version.

@@

After the Kiel Convention in 1989, the working group assisted
the International Phonetic Association in representing the IPA to the
\textsc{international organization for standardization} (ISO) and to the \textsc{text
encoding initiative} (TEI) \citep{EslingGaylord1993}. The working group's
formalization of the IPA, i.e.\ a full listing of agreed-upon computer codings
for phonetic symbols, was used in developing writing system descriptions, which
were at the time being solicited for scripts to be included in the new
multilingual international character encoding standards. The working group for
ISO/IEC 10646 and Unicode were two such initiatives.

@@

Character set ISO/IEC 10646 was approved by ISO, including the phonetic
characters submitted to them in May 1993. The set of IPA characters were
assigned UCS codes in 16-bit representation (in hexadecimal) and were published
as Tables 2 and 3 in \cite{EslingGaylord1993}, which include a graphical
representation of the IPA symbol, its IPA name, phonetic description, IPA
number, UCS code and AFII code.\footnote{The Association for Font
Information Interchange (AFII) was an international database of glyphs created
to promote the standardization of font data required to produce ISO/IEC 10646.} When the
character sets of ISO/IEC 10646 and the Unicode Standard later converged (see
Chapter~\ref{the-unicode-approach}), the IPA proposal was
included in the Unicode Standard Version 1.0 -- largely as we know it
today.

@@

It was a long journey, but the goal of achieving a single multilingual computing
environment has largely been accomplished. As such, we should not dismiss the IPA numbers 
or pre-Unicode encoding attempts, such as SAMPA/X-SAMPA, as misguided. The parallels 
between the IPA numbers and Unicode Code points, for example, are striking because both the IPA and 
the Unicode Consortium came up with the solution of an additional layer of indirection (an abstraction layer) 
between symbols/characters and encoding on the bit pattern level. SAMPA/X-SAMPA is also still useful 
as an input method for IPA in ASCII and required by some software.

@@

Unicode 88 provided the basic principles for the Unicode Standard's design -- 
pushing for 16-bit representations of characters with a clear distinction 
between characters and glyphs. Some of the contents of this status proposal 
of 1988 were reworked for inclusion in the early Unicode Standard pre-publication 
drafts and by August 1990, the proposal was in a (very) rough draft format. Its 
editors and the Unicode Working Group (the predecessor to the Unicode Technical 
Committee) worked together to lay out the proposed standard's structure and 
content. At this time, the proposal contained no code charts nor block descriptions.

### 2018-06-08 book/chapters/unicode_approach.tex
through paperhivers chapter 4

The conceptualization and terminology of writing systems was rejuvenated by
the development of the Unicode Standard, with major input from Mark Davis,
co-founder and long-term president of the Unicode Consortium. For many years,
``exotic'' writing systems and phonetic transcription systems on personal
computers were constrained by the American Standard Code for Information
Interchange (ASCII) character encoding scheme, based on the Latin script, which
only allowed for a strongly limited number of different symbols to be encoded.
This implied that users could either use and adopt the (extended) Latin alphabet
or they could assign new symbols to the small number of code points in the ASCII
encoding scheme to be rendered by a specifically designed font
\citep{BirdSimons2003}. In this situation, it was necessary to specify the font
together with each document to ensure the rightful display of its content. To
alleviate this problem of assigning different symbols to the same code points,
in the late 80's and early 90's the Unicode Consortium set itself the ambitious
goal of developing a single universal character encoding to provide a unique
number, a code point, for every character in the world's writing systems.
Nowadays, the Unicode Standard is the default encoding of the technologies that
support the World Wide Web and for all modern operating systems, software and
programming languages.

>>>>

The conceptualization and terminology of writing systems was rejuvenated by
the development of the Unicode Standard, with major input from Mark Davis,
co-founder and long-term president of the Unicode Consortium. For many years,
``exotic'' writing systems and phonetic transcription systems on personal
computers were constrained by the American Standard Code for Information
Interchange (ASCII) character encoding scheme, based on the Latin script, which
only allowed for a strongly limited number of different symbols to be encoded.
This implied that users could either use and adopt the (extended) Latin alphabet
or they could assign new symbols to the small number of code points in the ASCII
encoding scheme to be rendered by a specifically designed font
\citep{BirdSimons2003}. In this situation, it was necessary to specify the font
together with each document to ensure the rightful display of its content. To
alleviate this problem of assigning different symbols to the same code points,
in the late 80s and early 90s the Unicode Consortium set itself the ambitious
goal of developing a single universal character encoding to provide a unique
number, a code point, for every character in the world's writing systems.
Nowadays, the Unicode Standard is the default encoding of the technologies that
support the World Wide Web and for all modern operating systems, software and
programming languages.
